Generalizing the Model View Controller Paradigm

### 2006-10-26 inst/paper/MVC.tex
Changed the MVC paper.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@20825 bc3139a8-67e5-0310-9ffc-ced21a209358

\author{Elizabeth Whalen\\Harvard School of Public Health \\ 
        Robert Gentleman\\Fred Hutchinson Cancer Research Center}
\title{Generalizing the Model View Controller Paradigm}

@@

%% EAW: put new citations below
There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\citep{intGrUnwin, DynGraphics, IEEEVisual, GGobi, DataDesk}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset.  In some cases the
natural data representation of the subset is in a form other than that
of the initial data set.  This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a single, standard format.  We found that adding new views under such
conditions could be quite problematic and often led to very complex
code.  As we shall demonstrate, there are also issues that arise due to
mappings that are not one-to-one between elements on different views.
Hence we considered an approach that created new instances of
the MVC that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree.  Each parent can have multiple children, but the
children have a single parent.  Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow.  Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that can be linked in a natural way.  Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root). The parent MVC can choose to
react, or not, to messages from its children -- all control being in
the hands of the users, should they want it.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.  Also by interacting with the node with an arrow next to it in
Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
term, one can create a heatmap view that shows only the genes annotated at
that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
annotated at the signal transducer activity GO term have a blue label in
Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
This difference in these heatmaps indicates that views are synchronized based
on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
is the grandparent of Figure \ref{Fig:heatmapsub}.

@@

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
based more on component software constructed within an object oriented
programming environment.
%%FIXME: cite the JSS submission
We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} package of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
which we found results in a much shorter development time.

@@

%%FIXME: need to reference iPlots from Simon, and some of these refs need
%%to be put up at the front where I have a citations in parens
%%EAW: added a reference to iPlots below
Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as XLisp-Stat \cite{Lisp}, iPlots
\cite{iPlots}, and
GGobi \cite{GGobiMan}.  Implementing linked views is based on the
model-view-controller (MVC) design \cite{DesignPatterns}, which is a
widely used and well understood paradigm.

We take a fairly general view of interactivity. Users can interact via
a pointing device, such as a mouse, or via key strokes, or any of many
other types of input devices.  Interactivity means that there is a
response to some user action.  It is a powerful paradigm and allows
users to quickly navigate and explore complex documents or complex
data sets.  As an example, suppose that a user has two views: one view
is a graph where each node represents a particular set of genes while
the other view is a heatmap of gene expression data.  If the user
clicks on a node then the response to this action could be to
highlight all genes in the heatmap view that correspond to the
selected node.

@@

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view.  In the cartoon there are only two views, one of model A (note 
that once a data set is loaded in a MVC, we call it a model) and the
other of the linked model B.  When a user
interacts with the scatter plot view of model B, the response to
this interaction is for model B to update itself (this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because model B is linked
to model A, model B sends a message to model A.  In Step 4,
the information from model B is converted into information that
model A understands.  In Step 5, this information is received by
model A and model A updates itself.  Finally, in Step 6, model
A tells its view to update itself.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=6in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked models.  When a user interacts with
      the scatter plot, the first step, indicated by a circled 1,
      is to tell model B to update.  Then all views of model B must be
      updated.  Next any data that are linked to model B must be notified
      that model B changed.  This information from model B must be
      altered so that it can be understood by linked models, which is shown
      in the fourth step.  Finally, model A is notified that it should
      update in response to the change in model B and then all views of
      model A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

We felt that neither approach had much appeal.  A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.
We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a rectangular array. We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data. Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes. We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  But there are two distinct MVCs, not one. 
The city data have one MVC and the state data a second.  Even though 
there are two \Robject{MVC} objects here the views are linked.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own R software package.

@@

We support multiple loaded \Robject{MVCs}, but the control window menus
and the command line functions only perform operations on one
\Robject{MVC} object at a time.  Thus, we make use of the concept of
an \textit{active} \Robject{MVC} object, where the active
\Robject{MVC} object (and thus, the active model) means that
operations performed through the control window menus or through the
command line functions affect that \Robject{MVC} object. Users can
switch between active models, but since the interactions could be
quite different, there must always be an active MVC.

@@

Linking callback functions to events, such as a button click, a mouse
movement or a key press event, is what allows users to have an
interactive environment.  When an event occurs, the callback function
linked to the event determines the response to that event.  To create
a flexible environment it is important to be able to change the
response to an event.  For example, when a left button press event
occurs, a user may sometimes want a point to be colored and at other
times she may want a point to be hidden.  However, different responses
to an event may require functions with different signatures, where a function
signature is the combination of the function name along with its parameters.
In our example coloring a point requires two parameters, the new color to use
and the point to color, while hiding a point requires just one parameter, the
point to hide.

@@

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
%% EAW: When we decided that we wanted callback functions to be
%% interchangeable, we decided that the callback functions must have the same
%% signature so that calls to different callback functions would be
%% identical, as we wouldn't know ahead of time what function we would be
%% calling.  All callback functions must know which object in the plot to
%% act on, but there was no other universal information that all callback
%% functions needed.  Thus, for any other information that a callback function
%% would need, preprocessing functions were used.  As for that one parameter
%% that the callback function needed, the parameter type may be different
%% depending on the plot (and model) type - whatever is considered the object
%% of interest in a plot (and model).  Thus, for plots of data frames,
%% the object in the plot is considered a row from the data frame and for
%% plots of graphs, the object in the plot is considered a node in the graph. 
Since the intention is to be able to interchange the responses to an event,
the signatures of different callback functions should be the same.
In our design a callback function takes only one parameter. 
The class of that parameter depends on the type of model represented by the
\Robject{MVC}.  For example, for a graph model, the parameter corresponds to
the name of a node in the graph, while for a data frame model, the parameter
represents a case.  We expect a one-to-one relationship between the elements
on a view and the elements in the model.  Continuing the previous example,
each point on a scatter plot represents one case in the data frame model and
each node in the graph plot represents one node in the graph model.  Clearly
it is possible for two points in a scatter plot to overlap if they have the
same x axis and y axis values, but they are still considered two points
representing two cases in the model.

@@

%%FIXME: you use numbers, 1, 2, 3 and so on, for too many things. Use
%%letters to label objects, numbers are good for steps, as the numbers
%%mean something.
%% EAW: It's been changed.
An example of message passing within a \Robject{MVC} object is shown
in Figure~\ref{Fig:MPwithin}.  In this example, there are two views of the
model (view 1 and view 2), and the response to a user
performing a button click event over a point in the scatter plot is to
color the point red.  Setting the response to an event was discussed
in Section~\ref{Ssec:OneEvent}.  Here the user first clicks on a point
in view 1, which is shown as Step 1 (the circled 1 on the
figure).  In the figure, the point that has been clicked has an arrow
next to it.  Then, in Step 2, an update data message is sent to the
model, which updates itself accordingly, i.e. by changing the color
associated with the corresponding element.  After the model has been
updated, the views must be notified that a change has occurred so in
Step 3, an update view message is created and sent to all views of
this model.  In response to this message, the views update themselves in
Step 4 and now each view has one point that is colored red to reflect the
change in the model.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.5in, width=5in]{newMPwithin2.jpg}
    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatter plot (view 1).  Then an update data message is sent to
      the model in the second step, which alters the model.  Next in the third
      step all views of this model are sent an update view message, which
      causes the views to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

@@

Thinking of experimental data that are linked to meta-data as an
example, there are two \Robject{MVC} objects, one for the experimental
data and one for the meta-data.  In the example of experimental
microarray data that are linked to the Gene Ontology molecular
function graph (Section~\ref{sec:micro}), these two models are
linked in the following manner. The gene identifiers in the
experimental microarray data are linked to the GO terms.  Thus,
passing information between these two models requires that these
mappings be available.

@@

When a child \Robject{MVC} is created from an existing \Robject{MVC}, the
child model's \Rslot{linkData} slot is filled with the two functions,
\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in
the child \Robject{MVC} is updated, the \Rfunction{toParent} function in the
\Rslot{linkData} slot is used to convert the child model's data
change into useful information for the parent model.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=2.8in, width=4.5in]{newMP.jpg}
    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
      interacts with the y versus x scatter plot, the first step, indicated by
      a circled one, is to send an update data message to model B.  After
      model B has been updated, all views of this model must be notified, which
      is shown in the second step when an update view message is sent.  Next
      any data that are linked to model B must be notified that model B
      changed, which is shown in the third step as a send parent message.
      This message is converted into useful information for model A in
      the fourth step by the \Rfunction{toParent} function, and this
      information is sent to model A in the fifth step when an update data
      message is sent.  Finally, after model A is updated, an update view
      message is sent in the sixth step to update all views of model A. }
    \label{Fig:MessPass}
  \end{center}
\end{figure}

@@

The second goal of creating interactive views was met by using the
\Rpackage{RGtk} and \Rpackage{gtkDevice} packages to create devices that could
respond to events.  Using these packages made creating interactive views
particularly straightforward because any static R plot could be made
interactive by plotting it on a Gtk device and by having the plotting function
return information about the location of plot objects on the device.  In
addition the \Rclass{gEventFun} class was designed so that users could alter
the response to an event.  The response to an event could be as simple as
selecting the plot object that was just clicked by a mouse button or the
response could be as complicated as creating a new model.  This flexibility
lets the developer create a more varied interactive environment.

@@

\section*{Acknowledgements}
We are indebted to Dr. Duncan Temple Lang for the creation of the
\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
their use.

>>>>

\author{Elizabeth Whalen\\Harvard School of Public Health \\ 
        Duncan Temple Lang\\University of California at Davis \\
        Robert Gentleman\\Fred Hutchinson Cancer Research Center}
\title{Generalizing the Model View Controller Paradigm}

@@

%% EAW: put new citations below
There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\citep{intGrUnwin, DynGraphics, IEEEVisual, GGobi, DataDesk}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm, as introduced in the
Smalltalk system \cite{Smalltalk}, is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset.  In some cases the
natural data representation of the subset is in a form other than that
of the initial data set.  This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a single, standard format when implementing the design.  We found that 
adding new views under such conditions could be quite problematic and 
often led to very complex code.  As we shall demonstrate, there are 
also issues that arise due to mappings that are not one-to-one between 
elements on different views.  Hence we considered an approach that 
created new instances of the MVC that were linked to the parent that 
spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree.  Each parent can have multiple children, but the
children have a single parent.  Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow.  Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that can be linked in a natural way.  Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root).  Similarly, messages propagate 
to the child MVCs if the user initially interacts with a view in the 
parent MVC.  The MVC can choose to react, or not, to messages from its 
parent or children -- all control being in
the hands of the users, should they want it.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (though edges could also be considered the primary 
unit), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.  Also by interacting with the node with an arrow next to it in
Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
term, one can create a heatmap view that shows only the genes annotated at
that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
annotated at the signal transducer activity GO term have a blue label in
Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
This difference in these heatmaps indicates that views are synchronized based
on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
is the grandparent of Figure \ref{Fig:heatmapsub}.

@@

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
based more on component software constructed within an object oriented
programming environment.
%%FIXME: cite the JSS submission
We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} package of
D. Temple Lang \cite{RGtk}, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
which we found results in a much shorter development time.

@@

%%FIXME: need to reference iPlots from Simon, and some of these refs need
%%to be put up at the front where I have a citations in parens
%%EAW: added a reference to iPlots below
Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as DataDesk \cite{DataDesk},
XLisp-Stat \cite{Lisp}, iPlots \cite{iPlots}, and
GGobi \cite{GGobiMan}.  Implementing linked views is based on the
model-view-controller (MVC) design \cite{DesignPatterns}, which is a
widely used and well understood paradigm.

We take a fairly general view of interactivity.  Users can interact via
a pointing device, such as a mouse, or via key strokes, or any of many
other types of input devices.  Interactivity means that there is an
immediate response to some user action.  It is a powerful paradigm and allows
users to quickly navigate and explore complex documents or complex
data sets.  As an example, suppose that a user has two views: one view
is a graph where each node represents a particular set of genes while
the other view is a heatmap of gene expression data.  If the user
clicks on a node then the response to this action could be to
highlight all genes in the heatmap view that correspond to the
selected node.

@@

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view.  In the cartoon there are only two views, one of model A (note 
that once a data set is loaded in a MVC, we call it a model) and the
other of the linked model B.  When a user
interacts with the scatter plot view of model B, the response to
this interaction is for the controller to tell model B to update itself 
(this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because model B is linked
to model A, model B sends a message to model A.  In Step 4,
the information from model B is converted into information that
model A understands.  In Step 5, this information is received by
model A and model A updates itself.  Finally, in Step 6, model
A tells its view to update itself.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=6in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked models.  When a user interacts with
      the scatter plot, the first step, indicated by a circled 1,
      is for the controller to tell model B to update.  Then all views of
      model B must be updated.  Next any data that are linked to model B must
      be notified that model B changed.  This information from model B must be
      altered so that it can be understood by linked models, which is shown
      in the fourth step.  Finally, model A is notified that it should
      update in response to the change in model B and then all views of
      model A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

We felt that neither approach had much appeal.  A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.
We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a rectangular array. 
We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data.  Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes.  We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.  In our design, each 
MVC will contain one model and thus, linked models also indicate linked 
MVCs.

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  But there are two distinct MVCs, not one. 
The city data have one MVC and the state data a second.  Even though 
there are two \Robject{MVC} objects here, the views are linked.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own R software package.  More information on extending the classes can be
found in the \Rpackage{MVCClass} and \Rpackage{iSNetwork} Vignettes.

@@

We support multiple loaded \Robject{MVCs}, but the control window menus
and the command line functions only perform operations on one
\Robject{MVC} object at a time.  Thus, we make use of the concept of
an \textit{active} \Robject{MVC} object, where the active
\Robject{MVC} object (and thus, the active model) means that
operations performed through the control window menus or through the
command line functions affect that \Robject{MVC} object. Users can
switch between active models, but since the interactions could be
quite different, there must always be an active \Robject{MVC}.

@@

Linking callback functions to events, such as a button click, a mouse
movement or a key press event, is what allows users to have an
interactive environment.  When an event occurs, the callback function linked
to the event (this information is stored in the controller in the MVC)
determines the response to that event.  To create
a flexible environment it is important to be able to change the
response to an event.  For example, when a left button press event
occurs, a user may sometimes want a point to be colored and at other
times she may want a point to be hidden.  However, different responses
to an event may require functions with different signatures, where a function
signature is the combination of the function name along with its parameters.
In our example coloring a point requires two parameters, the new color to use
and the point to color, while hiding a point requires just one parameter, the
point to hide.

@@

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
%% EAW: When we decided that we wanted callback functions to be
%% interchangeable, we decided that the callback functions must have the same
%% signature so that calls to different callback functions would be
%% identical, as we wouldn't know ahead of time what function we would be
%% calling.  All callback functions must know which object in the plot to
%% act on, but there was no other universal information that all callback
%% functions needed.  Thus, for any other information that a callback function
%% would need, preprocessing functions were used.  As for that one parameter
%% that the callback function needed, the parameter type may be different
%% depending on the plot (and model) type - whatever is considered the object
%% of interest in a plot (and model).  Thus, for plots of data frames,
%% the object in the plot is considered a row from the data frame and for
%% plots of graphs, the object in the plot is considered a node in the graph.

Since the intention is to be able to interchange the responses to an event,
the signatures of different callback functions should be the same.
In our design a callback function takes only one parameter. 
The class of that parameter depends on the type of model represented by the
\Robject{MVC}.  For example, for a graph model, the parameter corresponds to
the name of a node in the graph, while for a data frame model, the parameter
represents a case.  We expect a one-to-one relationship between the elements
on a view and the elements in the model.  Continuing the previous example,
each point on a scatter plot represents one case in the data frame model and
each node in the graph plot represents one node in the graph model.  Clearly
it is possible for two points in a scatter plot to overlap if they have the
same x axis and y axis values, but they are still considered two points
representing two cases in the model.

@@

%%FIXME: you use numbers, 1, 2, 3 and so on, for too many things. Use
%%letters to label objects, numbers are good for steps, as the numbers
%%mean something.
%% EAW: It's been changed.
An example of message passing within a \Robject{MVC} object is shown
in Figure~\ref{Fig:MPwithin}.  In this example, there are two views of the
model (view 1 and view 2), and the response to a user
performing a button click event over a point in the scatter plot is to
color the point red.  Setting the response to an event was discussed
in Section~\ref{Ssec:OneEvent}.  Here the user first clicks on a point
in view 1, which is shown as Step 1 (the circled 1 on the
figure).  In the figure, the point that has been clicked has an arrow
next to it.  Then, in Step 2, an update data message is sent by the 
controller to the
model, which updates itself accordingly, i.e. by changing the color
associated with the corresponding element.  After the model has been
updated, the views must be notified that a change has occurred so in
Step 3, an update view message is created and sent to all views of
this model.  In response to this message, the views update themselves in
Step 4 and now each view has one point that is colored red to reflect the
change in the model.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.5in, width=5in]{newMPwithin2.jpg}
    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatter plot (view 1).  Then an update data message is sent by
      the controller to
      the model in the second step, which alters the model.  Next in the third
      step all views of this model are sent an update view message, which
      causes the views to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

@@

Thinking of experimental data that are linked to meta-data as an
example, there are two \Robject{MVC} objects, one for the experimental
data and one for the meta-data.  In the example of experimental
microarray data that are linked to the Gene Ontology molecular
function graph (as discussed in Section~\ref{sec:micro}), these two models 
are linked in the following manner. The gene identifiers in the
experimental microarray data are linked to the GO terms.  Thus,
passing information between these two models requires that these
mappings be available.

@@

When a child \Robject{MVC} is created from an existing \Robject{MVC}, the
child model's \Rslot{linkData} slot is filled with the two functions,
\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in
the child \Robject{MVC} is updated, the \Rfunction{toParent} function in the
\Rslot{linkData} slot is used to convert the child model's data
change into useful information for the parent model if there are any linked
elements in the parent model.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=2.8in, width=4.5in]{newMP.jpg}
    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
      interacts with the y versus x scatter plot, the first step, indicated by
      a circled one, is for the controller to send an update data message to
      model B.  After model B has been updated, all views of this model must
      be notified, which is shown in the second step when an update view
      message is sent.  Next any data that are linked to model B must be
      notified that model B changed, which is shown in the third step as a
      send parent message.  This message is converted into useful information
      for model A in the fourth step by the \Rfunction{toParent} function, and
      this information is sent to model A in the fifth step when an update data
      message is sent.  Finally, after model A is updated, an update view
      message is sent in the sixth step to update all views of model A. }
    \label{Fig:MessPass}
  \end{center}
\end{figure}

@@

The second goal of creating interactive views was met by using the
\Rpackage{RGtk} and \Rpackage{gtkDevice} packages \cite{RGtk} to create
devices that could respond to events.  Using these packages made creating
interactive views particularly straightforward because any static R plot could
be made interactive by plotting it on a Gtk device and by having the plotting
function return information about the location of plot objects on the device.
In addition the \Rclass{gEventFun} class was designed so that users could alter
the response to an event.  The response to an event could be as simple as
selecting the plot object that was just clicked by a mouse button or the
response could be as complicated as creating a new model.  This flexibility
lets the developer create a more varied interactive environment.

@@

%\section*{Acknowledgements}
%We are indebted to Dr. Duncan Temple Lang for the creation of the
%\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
%their use.
